---
title: "Analysis of Hotel Bookings"
author: "Isabel Ting"
date: "Februrary 9, 2023"
output:
  html_document: 
    toc: true
    toc_depth: 4
    css: style.css
---

```{r setup, include = F, echo = F}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### **I. Introduction**
The hotel bookings dataset used in this report was retrieved from Kaggle at https://www.kaggle.com/datasets/ahsan81/hotel-reservations-classification-dataset. 
This report aims to answer the question: How do you predict whether a guest will cancel or keep their hotel booking? I use logistic regression, two-sample t-tests, chi-square test of independence, and the random forest model to answer this question. 

Please see the Excel file in my GitHub repository for pivot tables / charts for basic trends of the data. 

### **II. Loading and Cleaning Data**

```{r}
# Load libraries
library(ggplot2)
library(dplyr)
library(randomForest)
library(tidymodels)
library(rsample)
library(tidyr)
library(caret)
library(Amelia)
library(knitr)
library(pander)
library(rempsyc)
library(effectsize)
library(readxl)
library(stargazer)

# Load dataset from excel
hotel <- read_excel("C:/Users/Isabel/OneDrive/Documents/Kaggle/Hotel Bookings/hotel-bookings/hotel.xlsx", 
                    sheet = 1)
```


```{r}
# Shorten/clean col names
hotel <- hotel %>% 
  rename(
    "booking_id"        = "Booking_ID", 
    "adults"            = "no_of_adults", 
    "kids"              = "no_of_children", 
    "weekend_nights"    = "no_of_weekend_nights",  
    "week_nights"       = "no_of_week_nights",
    "meal_plan"         = "type_of_meal_plan", 
    "parking_space"     = "required_car_parking_space",
    "room_type"         = "room_type_reserved", 
    "is_repeated_guest" = "repeated_guest",
    "prev_cancel"       = "no_of_previous_cancellations",
    "prev_kept"         = "no_of_previous_bookings_not_canceled",
    "room_price_usd"    = "avg_price_per_room",
    "special_req"       = "no_of_special_requests"
  )

# Change cols to factors
cols <- c("booking_id", "booking_status", "meal_plan", "room_type", 
          "arrival_year", "arrival_month", "is_repeated_guest")
hotel[cols] <- lapply(hotel[cols], as.factor)
# Note: 'arrival_date' not coded as a factor bc months can be categorized into seasons (non-linear)
# and years = distinct categories, but 'arrival_date' seems continuous 
# (beginning vs. end of month - linear)

# Convert euros to USD - use 1/1/2023 conversion rate
hotel$room_price_usd <- round((hotel$room_price_usd)*1.0703,2)

# Recode 'meal_plan' values for clarity
hotel$meal_plan <- recode(hotel$meal_plan, 
                          "Meal Plan 1"  = "breakfast only", 
                          "Meal Plan 2"  = "half set",  # breakfast & 1 other meal
                          "Meal Plan 3"  = "full set",  # all 3 meals
                          "Not Selected" = "none")
str(hotel)
```

#### **A. Summary**

After loading the hotel bookings dataset from Kaggle, I cleaned and shortened column names, changed categorical variables to factors, converted the room prices from euros to dollars, and recoded the values in the meal plan variable for clarity. 

### **III. Logistic Regression Model: Regress Booking Status on Predictors**

```{r}
# Set reference level of 'booking_status' to 'kept' 
hotel$booking_status <- relevel(hotel$booking_status, 
                                ref = "kept")

# Get random subset (1000 values) of 'hotel' data (too large for randomForest later on)
# Note: log reg & randomForest models use same training / test datasets
set.seed(123)
hotel_sample <- hotel[sample(1:nrow(hotel),1000),]

# Create training & test sets from 'hotel_sample'
set.seed(123)
hotel_split <- initial_split(hotel_sample, prop = 0.7)
hotel_train <- training(hotel_split)
hotel_test  <- testing(hotel_split)

# Logistic regression on training dataset: regress 'booking_status' on predictors
set.seed(123)
log_model_train <- glm(booking_status ~ meal_plan + btw_time + arrival_month + 
                         room_price_usd + special_req,
                         data = hotel_train, 
                         family = "binomial")

summary(log_model_train)
```


```{r, results = F}
# Get number of coefs in logistic regression model output
nrow(data.frame(log_model_train$coefficients))

# Get odds ratios for 'meal_planhalfset,' 'meal_plannone,' 'btw_time,' 'room_price_usd,' 'special_req'
or_half_meal   <- exp(log_model_train$coefficients[2])   
or_no_meal     <- exp(log_model_train$coefficients[3])  
or_btw_time    <- exp(log_model_train$coefficients[4])  
or_room_price  <- exp(log_model_train$coefficients[16]) 
or_special_req <- exp(log_model_train$coefficients[17]) 

# Get SDs for 5 predictors
meal_plan_half_sd <- summary(log_model_train)$coefficients[2, 2]
meal_plan_ns_sd   <- summary(log_model_train)$coefficients[3, 2]
btw_time_sd       <- summary(log_model_train)$coefficients[4, 2]
room_price_usd_sd <- summary(log_model_train)$coefficients[16,2]
special_req_sd    <- summary(log_model_train)$coefficients[17,2]

## Get 95% CI for 5 predictors
# meal_planhalf
upper_bound_meal_plan_half <- 
  exp((log_model_train$coefficients[2]) + 
        (1.96*meal_plan_half_sd))

lower_bound_meal_plan_half <- 
  exp((log_model_train$coefficients[2]) - 
        (1.96*meal_plan_half_sd))

# meal_plannone
upper_bound_meal_plan_ns <- 
  exp((log_model_train$coefficients[3]) + 
        (1.96*meal_plan_ns_sd))

lower_bound_meal_plan_ns <- 
  exp((log_model_train$coefficients[3]) - 
        (1.96*meal_plan_ns_sd))

# btw_time
upper_bound_btw_time <- 
  exp((log_model_train$coefficients[4]) + 
        (1.96*btw_time_sd))

lower_bound_btw_time <- 
  exp((log_model_train$coefficients[4]) - 
        (1.96*btw_time_sd))

# room_price_usd
upper_bound_room_price_usd <- 
  exp((log_model_train$coefficients[16]) + 
        (1.96*room_price_usd_sd))

lower_bound_room_price_usd <- 
  exp((log_model_train$coefficients[16]) - 
        (1.96*room_price_usd_sd))

# special_req
upper_bound_special_req <- 
  exp((log_model_train$coefficients[17]) + 
        (1.96*special_req_sd))

lower_bound_special_req <- 
  exp((log_model_train$coefficients[17]) - 
        (1.96*special_req_sd))
```


```{r}
# Print odds ratios
or_half_meal
or_no_meal   
or_btw_time    
or_room_price  
or_special_req 

# Print SDs
meal_plan_half_sd 
meal_plan_ns_sd   
btw_time_sd       
room_price_usd_sd 
special_req_sd 

# Print conf intervals
upper_bound_meal_plan_half
lower_bound_meal_plan_half 

upper_bound_meal_plan_ns 
lower_bound_meal_plan_ns 

upper_bound_btw_time 
lower_bound_btw_time 

upper_bound_room_price_usd 
lower_bound_room_price_usd 

upper_bound_special_req 
lower_bound_special_req 
```


```{r, results = F}
# Get probabilities of cancelling
pred <- log_model_train %>% 
  predict(hotel_test, 
          type = "response")

# Double-check probabilities are for cancelling booking 
# Note: probabilities refer to class dummy-coded as 1
contrasts(hotel_test$booking_status) # check good 

# Predict booking status for test dataset
pred_test <- ifelse(pred > 0.5, 
                    "cancelled", 
                    "kept")
```


```{r}
# Assess model accuracy
mean(pred_test == hotel_test$booking_status) # 81.3% accuracy
```

#### **A. Analysis**

Logistic regression is used to analyze the relationship between the response variable -- hotel booking status (kept or cancelled) -- and the following predictor variables: 

- Type of meal plan (breakfast only, half set, and full set),

- Number of days the room was booked in advance (*hereafter referred to as "between time"*), 

- Arrival month, 

- Room price (USD), and

- Number of special requests.  

The logistic regression model was built using the training subset (n = 700) of 1,000 observations randomly sampled from the original hotel bookings dataset (N = 36275). The model was used to predict booking statuses of the test subset (n = 300).  

I found that the predictors - selecting a half set meal plan (breakfast and one other meal) and not selecting a meal plan - were statistically significant at the 0.01 level, and the predictors - between time, room price, and number of special requests - were statistically significant at the 0.001 level. 

Furthermore, I found that, holding all other predictor variables constant: 

- The odds that the hotel booking was cancelled decreased by 30.6% (95% CI [.130, .719]) if the half set meal plan (breakfast and one other meal) were selected; 

- The odds that the hotel booking was cancelled increased by 139.2% (95% CI [.306, 3.379]) if a meal plan were not selected; 

- The odds that the hotel booking was cancelled increased by 1.9% (95% CI [.015, .022]) for every extra day the room was booked in advance;

- The odds that the hotel booking was cancelled increased by 2.0% (95% CI [.013, .028]) for every dollar increase in the room price; and

- The odds that the hotel booking was cancelled decreased by 28.4% (95% CI [.203, .397]) for every extra special request made by the guest. 

I tested the predictions of the logistic regression model on the test subset (n = 300) of the original hotel bookings dataset, and the model yielded an 81.3 percent accuracy rate, a relatively high accuracy rate. I will evaluate if this accuracy rate can be improved using the random forest model in Section V. 

### **IV. Two Sample T-Tests and Chi-Square Test of Independence**

```{r, results = F}
# Check for missing values
sum(is.na(hotel)) # no missing values

# Get two-sample t-test for 'btw_time' 
btw_time_test <- t.test(btw_time ~ booking_status, 
                        data = hotel, 
                        var.equal = F)

# Get two-sample t-test for 'room_price_usd'
price_test <- t.test(room_price_usd ~ booking_status, 
                     data = hotel, 
                     var.equal = F)

# Get two-sample t-test for 'special_req'
special_req_test <- t.test(special_req ~ booking_status, 
                           data = hotel, 
                           var.equal = F)
```


```{r}
# Print t-test results
btw_time_test
price_test
special_req_test
```


```{r, results = F}
# SD of 'btw_time' for cancelled bookings
sd_btw_time_c <- sd(hotel$btw_time[hotel$booking_status == "cancelled"]) # 98.95

# SD of 'btw_time' for kept bookings
sd_btw_time_k <- sd(hotel$btw_time[hotel$booking_status == "kept"]) # 64.03


# SD of 'room_price_usd' for cancelled bookings
sd_room_price_c <- sd(hotel$room_price_usd[hotel$booking_status == "cancelled"]) # 34.53

# SD of 'room_price_usd' for kept bookings
sd_room_price_k <- sd(hotel$room_price_usd[hotel$booking_status == "kept"]) # 38.39


# SD of 'special_req' for cancelled bookings
sd_special_req_c <- sd(hotel$special_req[hotel$booking_status == "cancelled"]) # 0.57

# SD of 'special_req' for kept bookings
sd_special_req_k <- sd(hotel$special_req[hotel$booking_status == "kept"]) # 0.84
```


```{r}
# Print SDs 
sd_btw_time_c
sd_btw_time_k

sd_room_price_c
sd_room_price_k

sd_special_req_c
sd_special_req_k
```


```{r}
# Get chi-square test of independence for 'meal_plan' & 'booking_status'
table(hotel$meal_plan, hotel$booking_status)

chisq.test(hotel$meal_plan, hotel$booking_status, 
           correct = F)
```

#### **A. Analysis**

Since the logistic regression model indicated that statistically significant predictors of booking status included meal plan, between time, room prices, and number of special requests, I ran two-sample t-tests on each of these latter three continuous predictors by booking status groups to further confirm these relationships. 

For **between time**, the results indicate that there is a statistically significant difference between the mean between times of guests who kept (M = 58.93, SD = 64.03) versus cancelled (M = 139.22, SD = 98.95) their hotel booking (t(16886) = 80.617, p < .001).

For **room prices**, the results indicate that there is a statistically significant difference between the mean room price of guests who kept (M = 106.96, SD = 38.39) versus cancelled (M = 118.36, SD = 34.53) their hotel booking (t(25929) = 28.45, p < .001).

For **number of special requests**, the results indicate that there is a statistically significant difference between the mean number of special requests of guests who kept (M = .76, SD = .83) versus cancelled (M = .33, SD = .57) their hotel booking (t(32323) = 56.425, p < .001).

Since meal plan is a categorical variable, I conducted a chi-square test of independence to examine the relationship between meal plan and booking status. The results showed that there is a significant relationship between the two variables, X^2(3, N = 36275) = 278.29, p < .001). Guests are less likely to cancel their booking if they selected a meal plan.   

### **V. Random Forest Model: Predicting Test Dataset of Hotel Bookings**

Finally, I used the machine learning algorithm, the random forest model, to predict the booking status of hotel guests. I predicted the booking statuses of guests in the test dataset (n = 700), a subset of the original hotel bookings dataset. 

This random forest model is appropriate for our classification task in predicting a *categorical* variable. (Note: The logistic regression conducted in Section III is still known as a regression task, not a categorical task, as logistic regressions produce continuous values of probabilities that are only later converted to Y = 0 or Y = 1 based on a threshold value.) 

Random forest models using the bootstrap aggregation (a.k.a. bagging) method along with a collection decision trees produce more accurate predictions for the response variable than basic decision trees. As such, I will tinker with the number of trees and number of variables randomly sampled as candidates at each split in a decision tree to produce the most accurate random forest model. Then, I will test my predictions against the test dataset (n = 300), a subset of the original dataset, to determine the accuracy of my model. 

```{r}
# Build random forest model to predict cancellation status
set.seed(123) 
rf_model <- randomForest(booking_status ~ meal_plan + btw_time + arrival_month + 
                           room_price_usd + special_req,
                         data = hotel_train, 
                         importance = T,
                         proximity = T)
rf_model # 14.57% OOB error rate

# Get df to plot error rates to see if 500 trees is enough
set.seed(123)
oob.error.hotel <- data.frame(
  Trees = rep(1:nrow(rf_model$err.rate), times = 3),
  Type  = rep(c("OOB", "Cancelled", "Kept"), 
              each = nrow(rf_model$err.rate)), 
  Error = c(rf_model$err.rate[,"OOB"], 
            rf_model$err.rate[,"cancelled"], 
            rf_model$err.rate[,"kept"])
)
# Note: Out-of-bag error rate: rate of wrongly classifying OOB sample (observations not chosen for the random samples); rF model still predicted values for OOB data

head(oob.error.hotel)

# Plot error rates
ggplot(data = oob.error.hotel, 
       aes(x = Trees, y = Error)) +
  geom_line(aes(color = Type)) + 
  ggtitle("Error Rates of Random Forest Model")
# Note: Higher error rate for 'cancelled' but 500 trees seems good 
# (stay with default ntree in rF model )
```


```{r}
# Determine if default mtry = 4 is best; test up to mtry = 10 
oob.values <- vector(length = 10)

# Create func to test diff rF models with diff mtry values
# (see which OOB error rate lowest)
for(i in 1:10){
  set.seed(123)
  temp.model    <- randomForest(booking_status ~ meal_plan + btw_time + arrival_month +
                                  room_price_usd + special_req,
                                data = hotel_train, 
                                mtry = i, 
                                ntree = 500)
  oob.values[i] <- temp.model$err.rate[nrow(temp.model$err.rate),1]
}
oob.values 
# mtry = 2 lowest OOB error rate (stay with default mtry)
```


```{r}
# Get plot of important variables in predicting booking status
set.seed(123)
varImpPlot(rf_model, 
           main = "Most Important Predictors of Booking Status") 
# Note: btw_time KEY predictor!
```

#### **A. Analysis: Plots of Most Important Predictors**

In both the Mean Decrease in Accuracy and Mean Decrease in Gini plots, between time ranks as the most important predictor in determining booking status by far, as shown through its outlier status on the far right of the plots. Although all the other predictors aside from arrival month have statistically significant coefficients in the logistic regression model, these plots suggest that between time has the strongest correlation with booking status among the predictors. 

```{r}
# Get prediction of cancellation statuses on test data using rF model
set.seed(123)
prediction <- predict(rf_model, hotel_test)
try1       <- data.frame(booking_id = hotel_test$booking_id, 
                         booking_status = prediction)

# Validate accuracy of prediction
print(confusionMatrix(prediction, hotel_test$booking_status))
# Note: 86% accuracy rate!
```

#### **B. Analysis: Accuracy Rate of Random Forest Model** 

An accuracy rate of 86% indicates that our random forest model accurately predicted the booking statuses for 86% of the data in the test hotel dataset. This is a higher accuracy rate than the logistic regression model that had an 81% accuracy rate. As such, I would prefer the random forest model to the logistic regression model in predicting hotel booking statuses. 

### **VI. Conclusion**

This report aimed to answer the question: How do you predict whether a guest will cancel or keep their hotel booking? According to our logistic regression, t-test, chi-square test of independence, and random forest model, **it appears that the most important variable in predicting booking status is the number of days in advance the booking was made (a.k.a. the time between booking date and arrival date)**. In general, the earlier in advance a guest books a hotel room, the more likely they will cancel it later on. This finding is logically sound - a guest's schedule may change as time passes, and they may no longer make their hotel booking. 