---
title: "Analysis of Hotel Bookings"
author: "Isabel Ting"
date: "Februrary 6, 2023"
output:
  html_document: 
    toc: true
    toc_depth: 4
    css: style.css
---

```{r setup, include = F, echo = F}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### **I. Introduction**
The hotel bookings dataset used in this report was retrieved from Kaggle at https://www.kaggle.com/datasets/ahsan81/hotel-reservations-classification-dataset. 
This report aims to answer the question: How does one predict whether a guest will cancel or keep their hotel booking? I use logistic regression, two-sample t-tests, and the random forest model to answer this question.

### **II. Loading and Cleaning Data**

```{r}
# Load data and libraries
hotel <- read.csv("hotel.csv", header = T, na.strings = c(""))

library(ggplot2)
library(dplyr)
library(randomForest)
library(tidymodels)
library(rsample)
library(tidyr)
library(caret)
library(Amelia)
library(knitr)
library(pander)
library(rempsyc)
library(effectsize)
```


```{r}
# Shorten/clean col names
hotel <- hotel %>% 
  rename(
    "booking_id"        = "Booking_ID", 
    "adults"            = "no_of_adults", 
    "kids"              = "no_of_children", 
    "weekend_nights"    = "no_of_weekend_nights",  
    "week_nights"       = "no_of_week_nights",
    "meal_plan"         = "type_of_meal_plan", 
    "parking_space"     = "required_car_parking_space",
    "room_type"         = "room_type_reserved", 
    "is_repeated_guest" = "repeated_guest",
    "prev_cancel"       = "no_of_previous_cancellations",
    "prev_kept"         = "no_of_previous_bookings_not_cancelled",
    "room_price_usd"    = "avg_price_per_room",
    "special_req"       = "no_of_special_requests"
  )

# Change cols to factors
cols <- c("booking_status", "meal_plan", "room_type", "arrival_year", 
          "arrival_month", "is_repeated_guest")
hotel[cols] <- lapply(hotel[cols], as.factor)
# Note: arrival_date is not coded as a factor since months can be categorized into seasons (non-linear)
# and years also appear to be distinct categories, but arrival_date seems it can be treated 
# as a continuous variable (linear - beginning vs. end of month)

# Convert euros to USD - use 1/1/2023 conversion rate
hotel$room_price_usd <- round((hotel$room_price_usd)*1.0703,2)
```

#### **A. Summary**

After loading the hotel dataset from Kaggle, I cleaned and shortened column names for clarity, changed categorical variables to factors, and converted the room prices from euros to dollars for clarity. 

### **III. Logistic Regression: Regress Booking Status on Predictors**

```{r}
# Set reference level of 'booking_status' to 'kept' 
# (important for logistic regression)
hotel$booking_status <- relevel(hotel$booking_status, 
                                ref = "kept")

# Get random subset (1000 values) of 'hotel' data (too large for randomForest later on)
# Note: Logistic regression and randomForest model will use the same training/test data sets
set.seed(123)
hotel_sample <- hotel[sample(1:nrow(hotel),1000),]

# Create training & test sets from 'hotel_sample'
set.seed(123)
hotel_split <- initial_split(hotel_sample, prop = 0.7)
hotel_train <- training(hotel_split)
hotel_test  <- testing(hotel_split)
```


```{r}
# Logistic regression on training set: regress 'booking_status' on predictors
log_model_train <- glm(booking_status ~ btw_time + arrival_month + 
                         arrival_date + room_type + prev_cancel + 
                         room_price_usd + special_req,
                       data = hotel_train, 
                       family = "binomial")

# Note: 'btw_time' = time between booking date & arrival date

summary(log_model_train)
```


```{r}
# Get number of coefs in logistic regression model output
nrow(data.frame(log_model_train$coefficients))

# Get odds ratios for 'btw_time,' 'room_price_usd,' 'special_req'
exp(log_model_train$coefficients[2])  # btw_time
exp(log_model_train$coefficients[20]) # room_price_usd
exp(log_model_train$coefficients[21]) # special_req

# Get SDs for 3 predictors
btw_time_sd       <- summary(log_model_train)$coefficients[2,2]
room_price_usd_sd <- summary(log_model_train)$coefficients[20,2]
special_req_sd    <- summary(log_model_train)$coefficients[21,2]

## Get 95% CI for 'btw_time,' 'room_price_usd,' 'special_req'
# btw_time
upper_bound_btw_time <- 
  exp(
  (log_model_train$coefficients[2]) + (1.96*btw_time_sd)
)
upper_bound_btw_time

lower_bound_btw_time <- 
  exp(
  (log_model_train$coefficients[2]) - (1.96*btw_time_sd)
)
lower_bound_btw_time

# room_price_usd
upper_bound_room_price_usd <- 
  exp(
  (log_model_train$coefficients[20]) + (1.96*room_price_usd_sd)
)
upper_bound_room_price_usd

lower_bound_room_price_usd <- 
  exp(
  (log_model_train$coefficients[20]) - (1.96*room_price_usd_sd)
)
lower_bound_room_price_usd

# special_req
upper_bound_special_req <- 
  exp(
  (log_model_train$coefficients[21]) + (1.96*special_req_sd)
)
upper_bound_special_req

lower_bound_special_req <- 
  exp(
  (log_model_train$coefficients[21]) - (1.96*special_req_sd)
)
lower_bound_special_req
```
#### **A. Analysis**

Logistic regression was used to analyze the relationship between the response variable -- hotel booking status -- and the predictor variables -- time between date of booking and date of arrival (**hereafter referred to as "between time"**), arrival month, arrival date, room type, number of previous booking cancellations, room price, and number of special requests. It was found that, holding all other predictor variables constant:

- The odds that the hotel booking was cancelled increased by 1.4% (95% CI [.011, .017]) for every extra day between the booking date and arrival date;
- The odds that the hotel booking was cancelled increased by 1.7% (95% CI [.009, .024]) for every dollar increase in the room price; and
- The odds that the hotel booking was cancelled decreased by 36.0% (95% CI [.262, .493]) for every extra special request made by the guest. 

Although *arrival month in December* was statistically significant at the p = .05 level, I will focus on the aforementioned predictors that were significant at the p = .001 level instead. 

### **IV. Two Sample T-Tests for Room Prices, Between Time, & Number of Special Requests**

```{r}
# Check for missing values
sum(is.na(hotel)) # no missing values

# Get two-sample t-test for 'btw_time' 
btw_time_test <- t.test(btw_time ~ booking_status, 
                        data = hotel, 
                        var.equal = F)

# Get two-sample t-test for 'room_price_usd'
price_test <- t.test(room_price_usd ~ booking_status, 
                     data = hotel, 
                     var.equal = F)

# Get two-sample t-test for 'special_req'
special_req_test <- t.test(special_req ~ booking_status, 
                           data = hotel, 
                           var.equal = F)
```


```{r}
btw_time_test
price_test
special_req_test
```


```{r}
# SD of 'btw_time' for cancelled bookings
sd(hotel$btw_time[hotel$booking_status == "cancelled"]) # 98.95

# SD of 'btw_time' for kept bookings
sd(hotel$btw_time[hotel$booking_status == "kept"]) # 64.03


# SD of 'room_price_usd' for cancelled bookings
sd(hotel$room_price_usd[hotel$booking_status == "cancelled"]) # 34.53

# SD of 'room_price_usd' for kept bookings
sd(hotel$room_price_usd[hotel$booking_status == "kept"]) # 38.39


# SD of 'special_req' for cancelled bookings
sd(hotel$special_req[hotel$booking_status == "cancelled"]) # 0.57

# SD of 'special_req' for kept bookings
sd(hotel$special_req[hotel$booking_status == "kept"]) # 0.84
```

#### **A. Analysis**

Since the logistic regression model indicated that statistically significant predictors of booking status included between time, room prices, and number of special requests, I decided to run two-sample t-tests on each of these predictors by the booking status groups to confirm these relationships. 

For **between time**, the results indicate that there is a statistically significant difference between the mean between times of guests who kept (M = 58.93, SD = 64.03) versus cancelled (M = 139.22, SD = 98.95) their hotel booking (t(16886) = 80.617, p < .001).

For **room prices**, the results indicate that there is a statistically significant difference between the mean room price of guests who kept (M = 106.96, SD = 38.39) versus cancelled (M = 118.36, SD = 34.53) their hotel booking (t(25929) = 28.45, p < .001).

For **number of special requests**, the results indicate that there is a statistically significant difference between the mean between times of guests who kept (M = .76, SD = .83) versus cancelled (M = .33, SD = .57) their hotel booking (t(32323) = 56.425, p < .001).

### **V. Random Forest Model: Predicting Test Dataset of Hotel Bookings**

Finally, I will use the machine learning algorithm, the random forest model, to produce predictions of the booking status of hotel guests. This model is appropriate for our classification task in predicting a categorical variable. (Note: The logistic regression conducted in Section III is still known as a regression task, not a categorical task, as logistic regressions produce continuous values of probabilities that are only later converted to Y = 0 or Y = 1 based on a threshold value.) 

Random forest models using the bootstrap aggregation (a.k.a. bagging) method along with a collection decision trees produce more accurate predictions for the response variable than basic decision trees. As such, I will tinker with the number of trees and number of variables randomly sampled as candidates at each split in a decision tree to produce the most accurate random forest model. Then, I will test my predictions against a subset of the hotel dataset to determine the accuracy of my model. 

```{r}
# Build random forest model 
# using every variable as predictor except booking_id 
set.seed(123) 
model <- randomForest(booking_status ~ . - booking_id,
                      data = hotel_train, 
                      ntree = 500, # kept at default
                      mtry = 2,    # better than default of 4
                      importance = T,
                      proximity = T)
model

# Plot error rates to see if 500 trees enough
set.seed(123)
oob.error.hotel <- data.frame(
  Trees = rep(1:nrow(model$err.rate), times = 3),
  Type  = rep(c("OOB", "Cancelled", "Kept"), 
              each = nrow(model$err.rate)), 
# Note: Out-of-bag error rate: rate of wrongly classifying OOB sample (observations not chose for the random samples); rF model still predicted values for OOB data
  Error = c(model$err.rate[,"OOB"], 
            model$err.rate[,"cancelled"], 
            model$err.rate[,"kept"])
)

ggplot(data = oob.error.hotel, 
       aes(x = Trees, y = Error)) +
  geom_line(aes(color = Type)) + 
  ggtitle("Error Rates of Random Forest Model")
# Note: Higher error rate for 'cancelled' but 500 trees seems good
```


```{r}
# Determine if default mtry = 4 is best; test up to mtry = 10 
oob.values <- vector(length = 10)

for(i in 1:10){
  set.seed(123)
  temp.model    <- randomForest(booking_status ~ . - booking_id, 
                                data = hotel_train, 
                                mtry = i, 
                                ntree = 500)
  oob.values[i] <- temp.model$err.rate[nrow(temp.model$err.rate),1]
}
oob.values 
# mtry = 2 lower OOB error rate 
# (w/o heavily increasing # of variables considered) - change in rF model
```


```{r}
# Get plot of important variables
set.seed(123)
varImpPlot(model, 
           main = "Most Important Predictors of Booking Status") 
# Note: btw_time KEY predictor!
```

#### **A. Analysis: Plots of Most Important Predictors**

In both the Mean Decrease in Accuracy and Mean Decrease in Gini plots, between time ranks as the most important predictor in determining booking status by far, as shown through its outlier status on the far right of the plot. This aligns with what I found in the logistic regression model, where between time was statistically significant at the p = .001 level. 

Further, room prices appears as the third most important predictor in both plots. This also aligns with the logistic regression results, where room prices was also statistically significant at the p = .001 level. 

```{r}
# Get prediction
set.seed(123)
prediction <- predict(model, hotel_test)
try1       <- data.frame(booking_id = hotel_test$booking_id, 
                         booking_status = prediction)

# Validate accuracy of prediction
print(confusionMatrix(prediction, hotel_test$booking_status))
# Note: 77% accuracy rate!
```

#### **B. Analysis: Accuracy Rate of Random Forest Model** 

An accuracy rate of 77% indicates that our random forest model accurately predicted the booking statuses for 79% of the data in the test hotel dataset.

### **VI. Conclusion**

This report aimed to answer the question: How does one predict whether a guest will cancel or keep their hotel booking? According to our logistic regression, t-test, and random forest model, **it appears that the most important variable in predicting booking status is the time between booking date and arrival date**. In general, the earlier in advance a guest books a hotel room, the more likely they will cancel it later on. This finding is logically sound - a guest's schedule may change as time passes, and they may no longer make their hotel booking. 

Room prices trails behind between time as a close second in importance of predictive value. The more expensive the hotel room, the higher the likelihood that the guest would cancel the booking.  
